Temperature and moisture effects on greenhouse gas emissions from deep active-layer boreal soils

### 2016-04-15 manuscript/cpcrw_manuscript.Rmd
Revisions to introduction, methods, results
Also moved MAX_MINCONC_TIME and MAX_MAXCONC_TIME definitions to
`0-functions.R`.

High latitude ecosystems are being subjected to rapid changes in climate {IPCC, 2013 #3916} and fire occurrence and intensity {Kasischke, 2010 #2631}. This has a wide variety of effects on ecosystems and society, notably in northwestern North America and Alaska {Hinzman, 2005 #4315}{Ju, 2016 #4320}. As a result, rising temperatures and increasing fire will result in permafrost degradation {Pastick, 2015 #4311}{Zhang, 2015 #4372} and changes in soil temperature and hydrology that are expected to increase soil greenhouse gas (GHG) fluxes to the atmosphere, promoting a positive climate feedback {Treat, 2015 #4217}{Koven, 2011 #3092}.

@@

Temperature and moisture typically have strong and often interactive influences on soil GHG emissions. A long literature of laboratory incubations, field observations, and meta-analyses has documented increased CO2 fluxes with rising temperature {Davidson, 2006 #1669}{Hashimoto, 2015 #4231}{Treat, 2015 #4217}. GHG responses to wetting and thawing dynamics are much less certain, with substantial variability between studies {Kim, 2012 #2863}. The anaerobic conditions common following permafrost thaw lower CO~2~ emissions but increase those of CH~4~ {Treat, 2015 #4217;Treat, 2014 #4213}, and interact with soil temperature effects in long-term incubations {Elberling, 2013 #3720}. Such laboratory studies have weaknesses {Podrebarac, 2016 #4365}, but their controlled environments provide an important way to elucidate the key mechanisms controlling GHG from high-latitude soils {Schuur, 2015 #4319}. Most studies have focused on surface soils or permafrost soils, however, neglecting deep active-layer soils that experience fluctuating temperature and, especially, moisture conditions. Such deep but unfrozen soils were identified as subject to strong effects from a two-decade warming experiment in the Alaskan Arctic {Sistla, 2013 #3666}.

The goal of this study was to examine how temperature and moisture control C loss and GHG (CO~2~ and CH~4~) emissions from soils sampled from the bottom of the annual active layer--i.e., directly above permafrost--in an Alaskan boreal forest. We also aimed to characterize the chemical and biological properties of these soils as they were subjected to a 100-day incubation at difference temperature and moisture levels. We hypothesized that (i) CO~2~ would be the dominant pathway for C loss in these largely aerobic soils; (ii) soils maintained at field moisture and high temperature would lose more C-CO~2~ than cores incubated at 4˚C, due to increased aerobic and anaerobic microbial activity; and (iii) core CH~4~ fluxes would be sensitive only to temperature.

@@

In the lab, the soil cores were stored at 4 °C for several days until they were weighed and prepared for incubation. At that point (11-12 August 2015), three fragmented or otherwise damaged cores were discarded, and the remaining cores were randomly assigned to one of six groups (N=6 in each group). These groups included two incubation temperatures (4 and 20 °C), each with two moisture treatments: one in which soil moisture was maintained at field conditions (~28% moisture by volume), and a drought treatment in which no water was added and cores were allowed to dry down to ~5% moisture by volume. The fifth group was a 20 °C "controlled drought" one, in which water was added so that these cores' moisture status would close matched those of the 4 °C "drought" cores, which we anticipated would dry more slowly than their 20 °C counterparts. The final 6-core group was used for destructive, pre-incubation measurements including moisture content, pH, soil carbon and nitrogen, and bulk density. Subsamples were collected and stored at -20 °C for dissolved organic carbon measurements.

@@

We measured the rise in GHG concentrations from its minimum value (in the first 10 seconds of measurement after the analyzer was sealed to the soil core) to its maximum (in the first 45 seconds), considering this as the flux rate in ppm s^-1^ (CO~2~) or ppb s^-1^ (CH~4~). This value was calculated for each sample, i.e. each individual core measurement throughout the 100-day incubation.

```{r compute_excludes, echo=FALSE}
# Compute outlier and exclusion numbers
# TODO: this isn't good--fragile if files move or change names
# Want everything in-memory (or something)
rn <- readLines("../outputs/3-remove/removed_data.csv") %>% length
sdn <- readLines("../outputs/summarydata.csv") %>% length
fluxn <- nrow(fluxdata)
```

@@

The GHG concentration measurements for each sample (i.e., each gas, core, and date/time) were used to calculate a linear rate of change (δc/δt), based on the concentration rise from a minimum (at 2-10 seconds after measurement began) to a maximum (at 10-45 seconds). Each core’s respiration flux (*F*) was then calculated following e.g. Steduto et al. {, 2002 #3383} as $F=\frac{\delta c}{\delta t} \frac{V}{M} \frac{P_a}{RT}$ where *V* is the core-specific system volume, *M* the core dry mass as determined at the end of the incubation, *P~a~* atmospheric pressure (101 kPa; the incubation chambers were ~120 m a.s.l.), *R* the universal gas constant (8.3 x 10^-3^ m^3^ kPa mol^-1^ K^-1^) and *T* the chamber air temperature (K) at time of measurement. The final respiration rate was expressed on a dry soil mass basis (mg C kg soil^-1^ day^-1^).

The effects of temperature, gravimetric water content, C percent, N percent, and DOC concentration on instantaneous gas fluxes were evaluated using a linear mixed-effects model fit by the R function *lme* in the R 'nlme' package, version `r packageVersion("nlme")`. Because the dependent variable (CO~2~ or CH~4~ flux) was non-normally distributed, it was transformed using a natural-logarithm (+`r FLUX_ADDITION` µg C g C^-1^ day^-1^ to ensure all positive fluxes) transformation. Core number was treated as a random effect in the model. We then used performed stepwise model selection by Akaike's information criterion (AIC) using the *stepAIC* function in the R 'MASS' package, version `r packageVersion("MASS")`. A linear mixed-effects model was also used to evaluate the effect of treatment on core water content.

@@

All data analysis and statistics were performed using R version `r R.version.string` {R Development Core Team, 2016 #4314}. This experiment was run as an "open experiment" with all analysis code, data (from raw instrument data to final summaries), diagnostics, etc., available at https://github.com/bpbond/cpcrw_incubation. The summarized flux data backing the main results have been archived at [DOI TO BE FILLED IN UPON ACCEPTANCE].

@@

Gas fluxes varied considerably over the course of the incubation (**Figures 2 and 3**). In the linear mixed-effects model, CO~2~ was strongly influenced by incubation chamber temperature, core gravimetric water content, and percent soil N (all P < 0.05, and the latter two P < 0.001; **Table 1**). (Note that percent C and percent N were highly correlated (r ≥ 0.98) for these cores. Because percent N was a slightlyl stronger predictor, it was retained in the model while percent C was excluded.) The interaction between water content and soil N was also highly significant (P < 0.001), with high-N cores having little relationship between water content and CO~2~ flux (data not shown). Methane fluxes were most strongly correlated with percent N, while water content exhibited significant interactions with percent N and DOC (**Table 2**). Neither temperature nor water content was a significant first-order predictor of CH~4~ fluxes, however.

@@

The cumulative flux numbers above result in CH~4~ temperature sensitivity (Q~10~) values of `r pn(Q10_cumulative["Field moisture", "CH4"], 1)` and `r pn(Q10_cumulative["Drought", "CH4"], 1)` for the field moisture and drought treatments, respectively; the corresponding Q~10~ values  based on cumulative CO~2~ were `r pn(Q10_cumulative["Field moisture", "CO2"], 1)` and `r pn(Q10_cumulative["Drought", "CO2"], 1)`. Computing Q~10~s based on fluxes normalized by water filled pore space decreased these values slightly: to `r pn(wfpsQ10_cumulative["Field moisture", "CH4"], 1)` and `r pn(wfpsQ10_cumulative["Drought", "CH4"], 1)` for CH~4~, respectively, and `r pn(wfpsQ10_cumulative["Field moisture", "CO2"], 1)` and `r pn(wfpsQ10_cumulative["Drought", "CO2"], 1)` for CO~2~.

@@

Several studies have measured microbial respiration and GHG fluxes from soils very close to our study site. Morishita et al. {, 2014 #4316} measured gas fluxes in the field at CPCRW and nearby forests, and found CO~2~ production to be correlated with both soil temperature and moisture, consistent with our results, TODO. Morishita et al. {, 2014 #4316} found however that CH~4~ uptake (they observed no emissions) was driven by temperature only; this is...

Waldrop et al. {, 2010 #4380} incubated  active-layer and permafrost soils from near Fairbanks, AK, under varied temperature and aerobic conditions, observing Q~10~ values of 9.0 (active layer) and 2.3 (permafrost); flux rates were 0.01-0.10 µmol CH~4~ day^-1^ g^-1^ () and 1-8 µg C-CO2 hr^-1^ g^-1^.

@@

Incubation experiments can only go so far though...

>>>>

High latitude ecosystems are being subjected to rapid changes in climate {IPCC, 2013 #3916} and fire occurrence and intensity {Kasischke, 2010 #2631}. This has a wide variety of effects on ecosystems and society, notably in northwestern North America and Alaska {Hinzman, 2005 #4315}{Ju, 2016 #4320}. In particular, rising temperatures and increasing fire will result in permafrost degradation {Pastick, 2015 #4311}{Zhang, 2015 #4372} and changes in soil temperature and hydrology that are expected to increase soil greenhouse gas (GHG) fluxes to the atmosphere, promoting a positive climate feedback {Treat, 2015 #4217}{Koven, 2011 #3092}.

@@

Temperature and moisture typically have strong and often interactive influences on soil GHG emissions. A long literature of laboratory incubations, field observations, and meta-analyses has documented increased CO2 fluxes with rising temperature {Davidson, 2006 #1669}{Hashimoto, 2015 #4231}{Treat, 2015 #4217}. GHG responses to wetting and thawing dynamics are much less certain, with substantial variability between studies {Kim, 2012 #2863}. The anaerobic conditions common following permafrost thaw are expected to lower CO~2~ emissions but increase those of CH~4~ {Treat, 2015 #4217;Treat, 2014 #4213}, and interact with soil temperature effects in long-term incubations {Elberling, 2013 #3720}. Decadal warming, and perhaps drying, trends in Alaska {Bieniek, 2014 #4389} will work to counteract these effects, with uncertain consequences for the carbon cycle and climate system.

The goal of this study was to examine how temperature and moisture control C loss and GHG (CO~2~ and CH~4~) emissions from soils sampled from the bottom of the annual active layer--i.e., directly above permafrost--in an Alaskan boreal forest. Most previous studies have focused on surface soils or permafrost soils, neglecting deep active-layer soils that experience fluctuating temperature and, especially, moisture conditions. Such deep but unfrozen soils were identified as subject to strong effects from a two-decade warming experiment in the Alaskan Arctic {Sistla, 2013 #3666}. We also aimed to characterize the chemical and biological properties of these soils as they were subjected to a 100-day incubation at difference temperature and moisture levels. We hypothesized that (i) CO~2~ would be the dominant pathway for C loss in these largely aerobic soils; (ii) soils maintained at field moisture and high temperature would lose more C-CO~2~ than cores incubated at 4˚C, due to increased aerobic and anaerobic microbial activity; and (iii) core CH~4~ fluxes would be sensitive only to temperature, as no anaerobic conditions were planned for the cores.

@@

In the lab, the soil cores were stored at 4 °C for several days until they were weighed and prepared for incubation. At that point (11-12 August 2015), three fragmented or otherwise damaged cores were discarded, and the remaining cores were randomly assigned to one of six groups (N=6 in each group). These groups included two incubation temperatures (4 and 20 °C), each with two moisture treatments: one in which soil moisture was maintained at field conditions (~28% moisture by volume), and a drought treatment in which no water was added and cores were allowed to dry down to ~5% moisture by volume. The fifth group was a 20 °C "controlled drought" one, in which water was added so that these cores' moisture status would close match those of the 4 °C "drought" cores, which we anticipated would dry more slowly than their 20 °C counterparts. The final 6-core group was used for destructive, pre-incubation measurements including moisture content, pH, soil carbon and nitrogen, and bulk density. Subsamples were collected and stored at -20 °C for dissolved organic carbon measurements.

@@

At each GHG measurement, we measured the rise in gas concentrations, considering this as the flux rate in ppm s^-1^ (CO~2~) or ppb s^-1^ (CH~4~). The GHG concentration measurements for each sample (i.e., each gas, core, and date/time) were used to calculate a linear rate of change (δc/δt), based on the concentration rise from a minimum (up to `r MAX_MINCONC_TIME` seconds after measurement began) to a maximum (at `r MAX_MINCONC_TIME`-`r MAX_MAXCONC_TIME` seconds). This value was calculated for each sample, i.e. each individual core measurement throughout the 100-day incubation.

```{r compute_excludes, echo=FALSE}
# Compute outlier and exclusion numbers
# TODO: this isn't good--fragile if files move or change names
rn <- readLines("../outputs/3-remove/removed_data.csv") %>% length
sdn <- readLines("../outputs/summarydata.csv") %>% length
fluxn <- nrow(fluxdata)
```

@@

Each core’s respiration flux (*F*) was calculated following e.g. Steduto et al. {, 2002 #3383} as $F=\frac{\delta c}{\delta t} \frac{V}{M} \frac{P_a}{RT}$ where *V* is the core-specific system volume, *M* the core dry mass as determined at the end of the incubation, *P~a~* atmospheric pressure (101 kPa; the incubation chambers were ~120 m a.s.l.), *R* the universal gas constant (8.3 x 10^-3^ m^3^ kPa mol^-1^ K^-1^) and *T* the chamber air temperature (K) at time of measurement. The final respiration rate was expressed on a dry soil mass basis (µg C g soil^-1^ day^-1^).

The effects of temperature, gravimetric water content, C percent, N percent, and DOC concentration on instantaneous gas fluxes were evaluated using a linear mixed-effects model fit by the R function *lme* in the R 'nlme' package, version `r packageVersion("nlme")`. Because the dependent variable (CO~2~ or CH~4~ flux) was non-normally distributed, it was transformed using a natural-logarithm (+`r FLUX_ADDITION` µg C g C^-1^ day^-1^ to ensure all positive fluxes, following Treat et al. 2015) transformation. Core number was treated as a random effect in the model. We then used performed stepwise model selection by Akaike's information criterion (AIC) using the *stepAIC* function in the R 'MASS' package, version `r packageVersion("MASS")`. A linear mixed-effects model was also used to evaluate the effect of treatment on core water content.

@@

All data analysis and statistics were performed using R version `r R.version.string` {R Development Core Team, 2016 #4314}. This experiment was run as an "open experiment" {Bond-Lamberty, 2016 #4388} with all analysis code, data (from raw instrument data to final summaries), diagnostics, etc., available at https://github.com/bpbond/cpcrw_incubation. The summarized flux data backing the main results have been archived at [DOI TO BE FILLED IN UPON ACCEPTANCE].

@@

These means conceal considerable variablility over the course of the incubation (**Figures 2 and 3**). In the linear mixed-effects model, CO~2~ was strongly influenced by incubation chamber temperature, core gravimetric water content, and percent soil N (all P < 0.05, and the latter two P < 0.001; **Table 1**). (Note that percent C and percent N were highly correlated (r ≥ 0.98) for these cores. Because percent N was a slightlyl stronger predictor, it was retained in the model while percent C was excluded.) The interaction between water content and soil N was also highly significant (`r pclean(summary(step_co2_lme)$tTable["WC_gravimetric:N_percent", "p-value"])`), with high-N cores having little relationship between water content and CO~2~ flux (data not shown). Methane fluxes were most strongly correlated with percent N, while water content exhibited significant interactions with percent N and DOC (**Table 2**). Neither temperature nor water content was a significant first-order predictor of CH~4~ fluxes, however.

@@

The cumulative flux numbers above result in CO~2~ temperature sensitivity (Q~10~) values of `r pn(Q10_cumulative["Field moisture", "CO2"], 1)` and `r pn(Q10_cumulative["Drought", "CO2"], 1)` for the field moisture and drought treatments, respectively; the corresponding Q~10~ values  based on cumulative CH~4~ were `r pn(Q10_cumulative["Field moisture", "CH4"], 1)` and `r pn(Q10_cumulative["Drought", "CH4"], 1)`. Computing Q~10~s based on fluxes normalized by water filled pore space changed these values only slightly: to `r pn(wfpsQ10_cumulative["Field moisture", "CO2"], 1)` and `r pn(wfpsQ10_cumulative["Drought", "CO2"], 1)` for CO~2~, respectively, and `r pn(wfpsQ10_cumulative["Field moisture", "CH4"], 1)` and `r pn(wfpsQ10_cumulative["Drought", "CH4"], 1)` for CH~4~.

@@

Several studies have measured microbial respiration and GHG fluxes from soils very close to our study site. Morishita et al. {, 2014 #4316} measured gas fluxes in the field at CPCRW and nearby forests, and found CO~2~ production to be correlated with both temperature and moisture, consistent with our results They found however that CH~4~ uptake (no emissions were observed) was driven by temperature only. Waldrop et al. {, 2010 #4380} incubated active-layer and permafrost soils from near Fairbanks, AK, under varied temperature and aerobic conditions, observing Q~10~ values of 9.0 (active layer) and 2.3 (permafrost) from -5 to 5 °C. Flux rates were 0.001-0.10 µmol CH~4~ day^-1^ g^-1^ (~`r pn(0.001 * 16 / 0.012 / 1000, 3)`-`r pn(0.1 * 16 / 0.012 / 1000, 3)` ng C g C^-1^ day^-1^), differing by orders of magnitude between sites but roughly similar to our observed emissions, and ~1-5 µg C-CO2 hr^-1^ g^-1^ (~`r pn(1 * 24 / 0.012, 0)`-`r pn(5 * 24 / 0.012, 0)` µg C g C^-1^ day^-1^), considerably higher than the CO~2~ rates observed from our cores.

@@

Incubation experiments can only go so far though...
Such laboratory studies have weaknesses {Podrebarac, 2016 #4365}, but their controlled environments provide an important way to elucidate the key mechanisms controlling GHG from high-latitude soils {Schuur, 2015 #4319}.
